Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-13 _paper.rmd
done with penguin case study

(ref:casepenguinsblfl) Checking what is learned from the cheem viewer. This is a plot of flipper length (`f_l`) and bill length (`b_l`), where the PI is highlighted by an asterisk. It is a Gentoo (purple) misclassified as a Chinstrap (orange). The PI has an unusually small `f_l` length which is the primary reason it is confused with a Chinstrap.
```{r casepenguinsblfl, echo=F, out.width="100%", fig.cap = "(ref:casepenguinsblfl)"}
knitr::include_graphics("./figures/case_penguins_BlFl.png")
```

There is more separation visible in the attribution space than the data space, as would be expected. The observed vs predicted plot reveals a handful of misclassified instances. The PI is a misclassified point (represented by the asterisk in the global view, and as a dashed vertical line in the tour view). The CI is a correctly classified point (represented by an $\times$ and a vertical dotted line).

The radial tour starts from the attribution projection of the misclassified instance (left plot). The important features identifed by SHAP in the (wrong) prediction for this instance is mostly `b_l` and `b_d` with small contributions of `f_l` and `wgt`. This projection is a view where the Gentoo (purple) looks much more likely for this instance than Chinstrap. For this instance we would like to know why a Gentoo has been wrongly labelled a Chinstrap. This combination of features is not particularly useful because the PI looks very much like other Gentoo penguins. To explore this we use the radial tour to vary the contribution of flipper length (`f_l`). The more that `f_l` contributes to the the projection the more this penguin looks like a Chinstrap. This suggests that `f_l` should have been an important feature for explaining the (wrong) prediction.

>>>>

(ref:casepenguinsblfl) Checking what is learned from the cheem viewer. This is a plot of flipper length (`f_l`) and bill length (`b_l`), where the PI is highlighted by an asterisk. It is a Gentoo (purple) misclassified as a Chinstrap (orange). The PI has an unusually small `f_l` length which is the primary reason it is confused with a Chinstrap.
```{r casepenguinsblfl, echo=F, out.width="60%", fig.cap = "(ref:casepenguinsblfl)"}
knitr::include_graphics("./figures/case_penguins_BlFl.png")
```

There is more separation visible in the attribution space than the data space, as would be expected. The observed vs predicted plot reveals a handful of misclassified instances. For this illustration we are going to explore why a Gentoo has been wrongly labelled as a Chinstrap.  The PI is a misclassified point (represented by the asterisk in the global view, and as a dashed vertical line in the tour view). The CI is a correctly classified point (represented by an $\times$ and a vertical dotted line).

The radial tour starts from the attribution projection of the misclassified instance (b, left). The important features identifed by SHAP in the (wrong) prediction for this instance is mostly `b_l` and `b_d` with small contributions of `f_l` and `wgt`. This projection is a view where the Gentoo (purple) looks much more likely for this instance than Chinstrap, that is, this combination of features is not particularly useful because the PI looks very much like other Gentoo penguins. To explore this we use the radial tour to vary the contribution of flipper length (`f_l`). (In our exploreation, this was the third variable explored. It is typically useful to explore the variables with larger contributions, here `b_l` and `b_d`, but when doing this nothing was revealed about how the PI differed from other Gentoos.) On varying `f_l` as it contributes more to the projection (b, right) we see that more, and more, this penguin looks like a Chinstrap. This suggests that `f_l` should be considered to be an important feature for explaining the (wrong) prediction.

Figure \@ref(fig:casepenguinsblfl) confirms that flipper length (`f_l`) is important for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted and we can see that the PI is closer to the Chinstrap group in these two variables, mostly because it has an unusually low value of flipper length relative to other Gentoos.
